---
title: "Midterm Project"
author: "ALEXANDER WILLIAMS (anw5ph)"

output: 
    html_document:
      number_sections: no
      toc: TRUE
      toc_depth: 4
      toc_float: true
      toc_collapsed: true
      theme: journal
      code_folding: show
---

```{r, echo=FALSE, results = 'hide', message=FALSE, warning=FALSE}

library(tidyverse)
library(ggplot2)

```

# Analyzing Trends Between a Student's Background and the Effect on Their Alcohol Consumption

## Introduction / Abstract

## Dataset Summary
```{r}
# Load the dataset

df1 <- read.csv('student-mat.csv')
df2 <- read.csv('student-por.csv')

head(df)

```

## Question 1

### Question
In this sample population, does alcohol consumption have an effect on number of absences in a math class or a Portuguese class. If so, does weekday or weekend alcohol consumption have more of an effect? 

### Prediction
We predict that alcohol consumption will have a positive correlation with the number of absences in both classes that a student has and that weekend alcohol consumption will have stronger positive trend than weekday alcohol consumption will have. 
### Chart & Chart Caption
```{r}

df1$Walc <- factor(df1$Walc, levels = c("1", "2", "3", "4", "5"), ordered = TRUE)
df1$Dalc <- factor(df1$Dalc, levels = c("1", "2", "3", "4", "5"), ordered = TRUE)
df1$health <- factor(df1$health, levels = c("1", "2", "3", "4", "5"), ordered = TRUE)

df1_sub_thirty <- subset(df1, absences <= 30)

ggplot(df1, aes(x = absences, fill = Dalc)) +
  geom_bar(position = position_dodge()) + 
  theme_bw()

ggplot(df1_sub_thirty, aes(x = absences, fill = Dalc)) +
  geom_bar(position = position_dodge()) + 
  theme_bw()

#ggplot(df1, aes(x = absences, y = Walc)) + 
#  geom_point() +
#  theme_bw()

```
### Discussion

### Chart Justification

## Question 2

### Question

### Prediction

### Chart & Chart Caption

### Discussion

### Chart Justification

## Conclusion

