---
title: "Midterm Project"
author: "ALEXANDER WILLIAMS (anw5ph)"

output: 
    html_document:
      number_sections: no
      toc: TRUE
      toc_depth: 4
      toc_float: true
      toc_collapsed: true
      theme: journal
      code_folding: show
---

```{r, echo=FALSE, results = 'hide', message=FALSE, warning=FALSE}

library(tidyverse)
library(ggplot2)

```

# Analyzing Trends Between a Student's Background and the Effect on Their Alcohol Consumption

## Introduction / Abstract

For the DS 2003 Midterm Project, we will be investigating the relationships of different variables in a dataset for Student Alcohol Consumption. This dataset was collected from two secondary schools in Portugal and contains a variety of variables ranging from demographics to drinking habits. We will use this dataset to to determine if there are any trends that we can identify regarding weekend drinking habits and frequency of going out for students, as well as, drinking habits and the final grade the student received in the course. As the data is from two schools in Portugal, the grading scale will be different and since the legal drinking age 18, there will be data with students drinking younger than 21.

## Dataset Summary

We will be using two datasets for this report. They both come from the same source and have the same columns (variables). The difference between the two is that one is for students in a Portuguese course and the other is for students in a math course. We will use the two to determine if the trends we observe for one course is also present in the other course.

For the first question of the report, we are using the following columns (variables): 
1. Walc - weekend alcohol consumption (factor: from 1 - very low to 5 - very high)
2. goout - going out with friends (factor: from 1 - very low to 5 - very high)

For the second question of the report, we are using the following columns (variables): 
1. 
2. 
3. 

Not used columns (variables):
1. school - student's school (binary: 'GP' - Gabriel Pereira or 'MS' - Mousinho da Silveira)
2. sex - student's sex (binary: 'F' - female or 'M' - male)
3. age - student's sex (binary: 'F' - female or 'M' - male)
4. address - student's home address type (binary: 'U' - urban or 'R' - rural)
5. famsize - family size (binary: 'LE3' - less or equal to 3 or 'GT3' - greater than 3)
6. Pstatus - parent's cohabitation status (binary: 'T' - living together or 'A' - apart)
7. Medu - mother's education (numeric: 0 - none, 1 - primary education (4th grade), 2 – 5th to 9th grade, 3 – secondary education or 4 – higher education)
8. Fedu - father's education (numeric: 0 - none, 1 - primary education (4th grade), 2 – 5th to 9th grade, 3 – secondary education or 4 – higher education)
9. Mjob - mother's job (nominal: 'teacher', 'health' care related, civil 'services' (e.g. administrative or police), 'at_home' or 'other')
10. Fjob - father's job (nominal: 'teacher', 'health' care related, civil 'services' (e.g. administrative or police), 'at_home' or 'other')
11. reason - reason to choose this school (nominal: close to 'home', school 'reputation', 'course' preference or 'other')
12. guardian - student's guardian (nominal: 'mother', 'father' or 'other')
13. traveltime - home to school travel time (numeric: 1 - <15 min., 2 - 15 to 30 min., 3 - 30 min. to 1 hour, or 4 - >1 hour)
14. studytime - weekly study time (numeric: 1 - <2 hours, 2 - 2 to 5 hours, 3 - 5 to 10 hours, or 4 - >10 hours)
15. failures - number of past class failures (numeric: n if 1<=n<3, else 4)
16. schoolsup - extra educational support (binary: yes or no)
17. famsup - family educational support (binary: yes or no)
18. paid - extra paid classes within the course subject (Math or Portuguese) (binary: yes or no)
19. activities - extra-curricular activities (binary: yes or no)
20. nursery - attended nursery school (binary: yes or no)
21. higher - wants to take higher education (binary: yes or no)
22. internet - Internet access at home (binary: yes or no)
23. romantic - Internet access at home (binary: yes or no)
24. famrel - quality of family relationships (numeric: from 1 - very bad to 5 - excellent)
25. freetime - free time after school (numeric: from 1 - very low to 5 - very high)
26. absences - number of school absences (numeric: from 0 to 93)
27. health - current health status (numeric: from 1 - very bad to 5 - very good)
28. G1 - first period grade (numeric: from 0 to 20)
29. G2 - second period grade (numeric: from 0 to 20)
30. G3 - final grade (numeric: from 0 to 20, output target)
31. Dalc - daily alcohol consumption (factor: from 1 - very low to 5 - very high)

```{r}
# Load the dataset

df1 <- read.csv('student-mat.csv')
df2 <- read.csv('student-por.csv')

head(df)

```

## Question 1

### Question
In this sample population, does alcohol consumption have an effect on number of absences in a math class or a Portuguese class. If so, does weekday or weekend alcohol consumption have more of an effect? 

### Prediction
We predict that alcohol consumption will have a positive correlation with the number of absences in both classes that a student has and that weekend alcohol consumption will have stronger positive trend than weekday alcohol consumption will have. 
### Chart & Chart Caption
```{r}

df1$Walc <- factor(df1$Walc, levels = c("1", "2", "3", "4", "5"), ordered = TRUE)
df1$Dalc <- factor(df1$Dalc, levels = c("1", "2", "3", "4", "5"), ordered = TRUE)
df1$health <- factor(df1$health, levels = c("1", "2", "3", "4", "5"), ordered = TRUE)

df1_sub_thirty <- subset(df1, absences <= 30)

ggplot(df1, aes(x = absences, fill = Dalc)) +
  geom_bar(position = position_dodge()) + 
  theme_bw()

ggplot(df1_sub_thirty, aes(x = absences, fill = Dalc)) +
  geom_bar(position = position_dodge()) + 
  theme_bw()

#ggplot(df1, aes(x = absences, y = Walc)) + 
#  geom_point() +
#  theme_bw()

```
### Discussion

### Chart Justification

## Question 2

### Question

### Prediction

### Chart & Chart Caption

### Discussion

### Chart Justification

## Conclusion

