---
title: "Untitled"
output: html_document
date: "2023-02-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---
title: "Midterm Project"
author: "ALEXANDER WILLIAMS (anw5ph)"

output: 
    html_document:
      number_sections: no
      toc: TRUE
      toc_depth: 4
      toc_float: true
      toc_collapsed: true
      theme: journal
      code_folding: show
---

```{r, echo=FALSE, results = 'hide', message=FALSE, warning=FALSE}

library(tidyverse)
library(ggplot2)
library(scales)

```

# Analyzing Trends Between a Student's Background and the Effect on Their Alcohol Consumption

## Introduction / Abstract

## Dataset Summary
```{r}
# Load the dataset
setwd("C:/Users/brian/OneDrive/Desktop/0- Spring 2023 S8/DS2003/Midterm Project")
df1 <- read.csv('student-mat.csv')
df2 <- read.csv('student-por.csv')

head(df1)

```

## Question 1

### Question
In this sample population, does alcohol consumption have an effect on number of absences in a math class or a Portuguese class. If so, does weekday or weekend alcohol consumption have more of an effect? 

### Prediction
We predict that alcohol consumption will have a positive correlation with the number of absences in both classes that a student has and that weekend alcohol consumption will have stronger positive trend than weekday alcohol consumption will have. 
### Chart & Chart Caption
```{r}

df1$Walc <- factor(df1$Walc, levels = c("1", "2", "3", "4", "5"), ordered = TRUE)
df1$Dalc <- factor(df1$Dalc, levels = c("1", "2", "3", "4", "5"), ordered = TRUE)
df1$health <- factor(df1$health, levels = c("1", "2", "3", "4", "5"), ordered = TRUE)

df1_sub_thirty <- subset(df1, absences <= 30)

ggplot(df1, aes(x = absences, fill = Dalc)) +
  geom_bar(position = position_dodge()) + 
  theme_bw()

ggplot(df1_sub_thirty, aes(x = absences, fill = Dalc)) +
  geom_bar(position = position_dodge()) + 
  theme_bw()

#ggplot(df1, aes(x = absences, y = Walc)) + 
#  geom_point() +
#  theme_bw()

```
### Discussion

### Chart Justification

## Question 2

### Question
Does going out at a higher frequency negatively affect students' final grade in the math class as opposed to the students who responded that they go out less frequently?
### Prediction
We think that the students who go out more receive worse grades than their peers who go out less. This means that going out will have a negative correlation with final grades, as we predict that students who respond to their level of going out with a 1 receive higher grades in the class on average.
### Chart & Chart Caption
```{r}
# Converting both G3 and goout into factors
df1$G3 <- factor(df1$G3, levels=c("0", "1", "2", "3","4", "5", "6", "7", "8", "9", "10",
                                  "11", "12", "13", "14", "15", "16", "17", "18", "19", "20"), 
                 ordered= TRUE)
df1$goout <- factor(df1$goout, levels=c("1", "2", "3", "4", "5"), ordered=TRUE)


# Subsetting df1 by grade intervals
### A: G3=17-20
### B: G3=13-16
### C: G3=9-12
### D: G3=5-8
### F: G3=0-4
df1G3_1720 <- subset(df1, G3 >= 17)

df1G3_1316 <- subset(df1, G3>=13 & G3<=16)

df1G3_912 <- subset(df1, G3>=9 & G3<=12)

df1G3_58 <- subset(df1, G3>=5 & G3<=8)

df1G3_04 <- subset(df1, G3>=0 & G3<=4)


# Creating histograms for each grade interval
### Each histogram is measured by the percentage of students that received
### a grade in its specified G3 interval based on their response to
### how often they go out on a scale from 1-5.
###### i.e. At what level (1-5) of going out did each student list
######      and what grade did they receive.
As_hist <- ggplot(df1G3_1720, aes(x=goout, y=..count.., fill=goout)) +
  geom_bar(aes(y = (..count..)/sum(..count..))) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1L)) +
  xlab(label='Measure of Going Out') +
  ylab(label='Percentage') +
  labs(title='Percentage of Students by Level of Going Out',
       subtitle='"A" students with a final grade of 17-20')

Bs_hist <- ggplot(df1G3_1316, aes(x=goout, y=..count.., fill=goout)) +
  geom_bar(aes(y = (..count..)/sum(..count..))) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1L)) +
  xlab(label='Measure of Going Out') +
  ylab(label='Percentage') +
  labs(title='Percentage of Students by Level of Going Out',
       subtitle='"B" students with a final grade of 13-16')

Cs_hist <- ggplot(df1G3_912, aes(x=goout, y=..count.., fill=goout)) +
  geom_bar(aes(y = (..count..)/sum(..count..))) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1L)) +
  xlab(label='Measure of Going Out') +
  ylab(label='Percentage') +
  labs(title='Percentage of Students by Level of Going Out',
       subtitle='"C" students with a final grade of 9-12')

Ds_hist <- ggplot(df1G3_58, aes(x=goout, y=..count.., fill=goout)) +
  geom_bar(aes(y = (..count..)/sum(..count..))) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1L)) +
  xlab(label='Measure of Going Out') +
  ylab(label='Percentage') +
  labs(title='Percentage of Students by Level of Going Out',
       subtitle='"D" students with a final grade of 5-8')

Fs_hist <- ggplot(df1G3_04, aes(x=goout, y=..count.., fill=goout)) +
  geom_bar(aes(y = (..count..)/sum(..count..))) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1L)) +
  xlab(label='Measure of Going Out') +
  ylab(label='Percentage') +
  labs(title='Percentage of Students by Level of Going Out',
       subtitle='"F" students with a final grade of 0-4')

grades_by_goout <- list(As_hist, Bs_hist, Cs_hist, Ds_hist, Fs_hist)
grades_by_goout

```

### Discussion

### Chart Justification

## Conclusion

```{r}
df1G3_q1 <- subset(df1, G3 <=8)

df1G3_q2 <- subset(df1, G3>=9 & G3<=10)

df1G3_q3 <- subset(df1, G3>=11 & G3<=13)

df1G3_q4 <- subset(df1, G3>=14 & G3<=20)

q1_hist <- ggplot(df1G3_q1, aes(x=goout, y=..count.., fill=goout)) +
  geom_bar(aes(y = (..count..)/sum(..count..))) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1L)) +
  xlab(label='Measure of Going Out') +
  ylab(label='Percentage') +
  labs(title='Percentage of Students by Level of Going Out',
       subtitle='Students in the first quartile (grade of 1-8)')

q2_hist <- ggplot(df1G3_q2, aes(x=goout, y=..count.., fill=goout)) +
  geom_bar(aes(y = (..count..)/sum(..count..))) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1L)) +
  xlab(label='Measure of Going Out') +
  ylab(label='Percentage') +
  labs(title='Percentage of Students by Level of Going Out',
       subtitle='Students in the second quartile (grade of 9-10)')

q3_hist <- ggplot(df1G3_q3, aes(x=goout, y=..count.., fill=goout)) +
  geom_bar(aes(y = (..count..)/sum(..count..))) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1L)) +
  xlab(label='Measure of Going Out') +
  ylab(label='Percentage') +
  labs(title='Percentage of Students by Level of Going Out',
       subtitle='Students in the third quartile (grade of 11-13)')

q4_hist <- ggplot(df1G3_q4, aes(x=goout, y=..count.., fill=goout)) +
  geom_bar(aes(y = (..count..)/sum(..count..))) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1L)) +
  xlab(label='Measure of Going Out') +
  ylab(label='Percentage') +
  labs(title='Percentage of Students by Level of Going Out',
       subtitle='Students in the fourth quartile (grade of 14-20)')

gradequartile_by_goout <- list(q1_hist, q2_hist, q3_hist, q4_hist)
gradequartile_by_goout

```



```{r}
third1 <- ggplot(df1[which(df1$G3>=0 & df1$G3<=9),], aes(x=goout, y=..count.., fill=goout)) +
  geom_bar(aes(y = (..count..)/sum(..count..))) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1L)) +
  xlab(label='Measure of Going Out') +
  ylab(label='Percentage') +
  labs(title='Percentage of Students by Level of Going Out',
       subtitle='Students in the first third (grade of 0-9)')

third2 <- ggplot(df1[which(df1$G3>=10 & df1$G3<=12),], aes(x=goout, y=..count.., fill=goout)) +
  geom_bar(aes(y = (..count..)/sum(..count..))) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1L)) +
  xlab(label='Measure of Going Out') +
  ylab(label='Percentage') +
  labs(title='Percentage of Students by Level of Going Out',
       subtitle='Students in the first third (grade of 10-12)')

third3 <- ggplot(df1[which(df1$G3>=13 & df1$G3<=20),], aes(x=goout, y=..count.., fill=goout)) +
  geom_bar(aes(y = (..count..)/sum(..count..))) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1L)) +
  xlab(label='Measure of Going Out') +
  ylab(label='Percentage') +
  labs(title='Percentage of Students by Level of Going Out',
       subtitle='Students in the first third (grade of 13-20)')

list(third1, third2, third3)

```
